One-stage point transect distance sampling using iterated integrated nested Laplace approximations

### 2020-11-09 paper-v2.tex
Rest of Finn comments except posterior N and Cov(exp(grf))

To evaluate the integral we use polar coordinates notation $s_k(r, \theta) = s_k + r\left[\cos\theta, \sin\theta \right]^T$ to represent locations in each sampling unit $\Omega_k$.   In general, the thinning function depends only on $r$ and not on $\theta$.  We assume this is the case in the following and use the shorthand $g_k(r) = g(s_k(r, \theta))$. The integral in the likelihood can be simplified using an assumption that $\lambda(s)$ is linear within each sampling unit.  This assumption implies that $\lambda(s_k(r, \theta)) + \lambda(s_k(r, \theta + \pi)) = 2\lambda(s_k)$ and so
\begin{align*}
	\int_0^{2\pi} \lambda(s(r, \theta))g_k(r)\mathrm{d}\theta &= \int_0^\pi \left[\lambda(s(r, \theta)) + \lambda(s(r, \theta + \pi)) \right] g_k(r)\mathrm{d}\theta \\
	&= 2\pi \lambda(s_k)g(r)
\end{align*}
We note that this is a relaxation of the traditional assumption in distance sampling that the intensity is constant within each sampling unit.  A similar relaxation is also possible in the case of line transects \citep{yuan_point_2017} and this assumption complements the choice of piece-wise linear basis functions we use to model $\lambda$ (see section 5).  Integrating with respect to $r$ and accounting for the change of variables gives $\int \lambda(s)g(s) \mathrm{d}s = \sum_{k=1}^K 2\pi \lambda(s_k) \int_0^W r g_k(r)\mathrm{d}r$.   For each sampling unit we approximate the one-dimensional integral with respect to $r$ using a midpoint integration method with $M$ integration locations $r_{k1}, \ldots, r_{kM}$ and associated weights $\alpha_{k1}, \ldots, \alpha_{kM}$.  This gives
\begin{equation*}
	\int\displaylimits_{s \in \Omega} \lambda(s)g(s)\mathrm{d}s \approx \sum_{k=1}^K \sum_{j=1}^M \tilde{\alpha}_{kj} \tl(s_{kj}) \;\;\; ,
\end{equation*}
where $\tilde{\alpha}_{kj} = 2\pi \alpha_{kj}r_{kj}$ and $\tl(s_{kj}) = \lambda(s_k) g_k(r_{kj})$.

@@

We use a GMRF representation of $\xi(s)$ following the approach of \cite{lindgren_explicit_2011} based on specifying $\xi(s)$ via a SPDE.  Given a finite element mesh on the spatial domain with $L$ nodes and associated piece-wise linear basis functions $\phi_1, \ldots, \phi_L$ we represent the GRF as $\xi(s) = \sum_l \xi_l \phi_l(s)$.  The parameters $\xi_1, \ldots, \xi_L$ form a GMRF with sparse prior precision matrix $\bm{Q}_{\xi} = \frac{1}{\tau^2}\left(\kappa^4\bm{C} + 2\kappa^2\bm{G_1} + \bm{G_2}\right)$ where $\bm{C}$, $\bm{G_1}$, $\bm{G_2}$ are all sparse matrices the elements of which are defined via an approximate solution to a SPDE (see Supplementary material). The parameters $\tau$ and $\kappa$ control the shape and rate of decay of the Mat\'ern covariance function.

The SPDE approach can at first appear to be fundamentally different to other spatial random effect approaches such as penalised regression splines.  However, the two methods are closely related and the GRF model component in our model plays a similar role to a tensor product spline on $s$ with an appropriate smoothing penalty.  See \cite{yue_bayesian_2014, miller_understanding_2019} for more details on the similarites between penalised regression splines and the SPDE approach.

@@

The predicted mean of the posterior intensity is shown in \autoref{fig:intensity-mean-cv} along with a map of the coefficient of variation (CV) of the posterior intensity field on a regular prediction grid with cell area of approximately 1.7 hectares.  
\begin{figure}[h]
	\begin{center}
		\includegraphics[scale=0.525]{figures/intensity_mean_cv_sd.png}
		\caption{The A) mean, B) CV and C) SD of the posterior intensity field}
		\label{fig:intensity-mean-cv}
	\end{center}
\end{figure}
This shows a region of high intensity in the south and much lower intensity in the north, agreeing with a standard two-stage analysis of the same data (see Supplementary material).  The CV plot shows that the CV in the posterior intensity field is lower in areas with greater sampling effort.  There is a clear indication of preferential sampling with more survey effort in the south, where densities are higher, compared to the north  (see \autoref{fig:2002studyareapointspt}).  In the north, estimated intensities are lower, which also contributes to larger CV values.  The standard deviation map shows stronger overall posterior variability in the south where intensity is larger.  We also map the lower and upper quantiles corresponding to a 95\% credible interval for each prediction location.  [[Say something about these - hard to know what to say about them?]]   (\autoref{fig:intensity-quantiles}).
\begin{figure}[h]
	\begin{center}
		\includegraphics[scale=0.525]{figures/intensity_quantiles.png}
		\caption{The predicted posterior intensity quantiles:  A) 0.025 quantile B) 0.975 quantile.  Both plots plots use the same colour scale}
		\label{fig:intensity-quantiles}
	\end{center}
\end{figure}

@@

We plot the posterior abundance estimate using a Monte Carlo method.
\begin{figure}[h]
	\begin{center}
		\includegraphics[scale=0.525]{figures/realized_abundance_posterior.png}
		\caption{Posterior abundance}
		\label{fig:realized-abundance-posterior}
	\end{center}
\end{figure}
Let $n$ denote the abundance of the population within the region of interest and $N = \int_{\Omega}\lambda(s)\mathrm{d}s$ its corresponding random variable within our Bayesian framework.  Integrating the mean of the posterior intensity field provides a point estimate for the abundance, i.e. the expected abundance.  We approximate the posterior $\pi(N | \lambda)$ by a Monte Carlo method.  Taking $m$ Monte Carlo samples  $\lambda_1, \ldots, \lambda_m$ of the posterior intensity field we estimate the posterior for the abundance as $\pi(N | \lambda) \approx 1 / m \sum_{i=1}^m \pi (N | \lambda = \lambda_i)$. Each $\pi(N | \lambda = \lambda_i)$ is a Poisson probability mass function with rate parameter $\int_{\Omega}\lambda_i(s)\mathrm{d}s$. \autoref{fig:realized-abundance-posterior} shows the approximate posterior for $N$ with $m = 20,000$.  This allows us to estimate the probability for any specific value of realised abundance $n$ by calculating $\pi(N = n | \lambda)$.

@@

There are several natural extensions to the model.  For this analysis we only considered a single survey year.  Spatio-temporal extensions to this are possible, for example by considering a space-time interaction models.  \texttt{R-INLA} has the ability to fit space-time effects that can be represented as a Kronecker product of the relevant precision matrices (e.g. \citep{yuan_point_2017}).  An example of this type of space-time interaction effect is an interaction between the spatial SPDE with an auto-regressive temporal process.  \cite{blangiardo_spatial_2013} show how this can be done with numerous examples and code snippets.

>>>>

To evaluate the integral we use polar coordinates notation $s_k(r, \theta) = s_k + r\left[\cos\theta, \sin\theta \right]^T$ to represent locations in each sampling unit $\Omega_k$.   In general, the thinning function depends only on $r$ and not on $\theta$.  We assume this is the case in the following and use the shorthand $g_k(r) = g(s_k(r, \theta))$. The integral in the likelihood can be simplified using an assumption that $\lambda(s)$ is linear within each sampling unit.  This assumption implies that $\lambda(s_k(r, \theta)) + \lambda(s_k(r, \theta + \pi)) = 2\lambda(s_k)$ and so
\begin{align*}
	\int_0^{2\pi} \lambda(s_k(r, \theta))g_k(r)\mathrm{d}\theta &= \int_0^\pi \left[\lambda(s_k(r, \theta)) + \lambda(s_k(r, \theta + \pi)) \right] g_k(r)\mathrm{d}\theta \\
	&= 2\pi \lambda(s_k)g_k(r)
\end{align*}
We note that this is a relaxation of the traditional assumption in distance sampling that the intensity is constant within each sampling unit.  A similar relaxation is also possible in the case of line transects \citep{yuan_point_2017} and this assumption complements the choice of piece-wise linear basis functions we use to model $\lambda$ (see section 5).  Integrating with respect to $r$ and accounting for the change of variables gives $\int \lambda(s)g(s) \mathrm{d}s = \sum_{k=1}^K 2\pi \lambda(s_k) \int_0^W r g_k(r)\mathrm{d}r$.   For each sampling unit we approximate the one-dimensional integral with respect to $r$ using a midpoint integration method with $M$ integration locations $r_{k1}, \ldots, r_{kM}$ and associated weights $\alpha_{k1}, \ldots, \alpha_{kM}$.  This gives
\begin{equation*}
	\int\displaylimits_{s \in \Omega} \lambda(s)g(s)\mathrm{d}s \approx \sum_{k=1}^K \sum_{j=1}^M \tilde{\alpha}_{kj} \tl(s_{kj}) \;\;\; ,
\end{equation*}
where $\tilde{\alpha}_{kj} = 2\pi \alpha_{kj}r_{kj}$ and $\tl(s_{kj}) = \lambda(s_k) g_k(r_{kj})$.

@@

We use a GMRF representation of $\xi(s)$ following the approach of \cite{lindgren_explicit_2011} based on specifying $\xi(s)$ via a SPDE.  Given a finite element mesh on the spatial domain with $L$ nodes and associated piece-wise linear basis functions $\phi_1, \ldots, \phi_L$ we represent the GRF as $\xi(s) = \sum_l \xi_l \phi_l(s)$.  The parameters $\xi_1, \ldots, \xi_L$ form a GMRF with sparse prior precision matrix $\bm{Q}_{\xi} = \frac{1}{\tau^2}\left(\kappa^4\bm{C} + 2\kappa^2\bm{G_1} + \bm{G_2}\right)$ where $\bm{C}$, $\bm{G_1}$, $\bm{G_2}$ are all sparse matrices the elements of which are defined via an approximate solution to a SPDE (see Supplementary material). The parameters $\tau$ and $\kappa$ control the shape and rate of decay of the Mat\'ern covariance function. In order to specify priors on the Mat\'ern covariance below we use a reparameterisation to a range and variance parameter.  When $\nu = 1$ and the domain is of dimension two the reparameterisation is $\rho = \sqrt{8} / \kappa$ and $\sigma^2 = 1 / (4\pi\kappa^2\tau^2)$ \citep{blangiardo_spatial_2013}.

The SPDE approach can at first appear to be fundamentally different to other spatial random effect approaches such as penalised regression splines.  However, the two methods are closely related and the GRF model component in our model plays a similar role to a tensor product spline with an appropriate smoothing penalty.  See \cite{yue_bayesian_2014, miller_understanding_2019} for more details on the similarities and differences between penalised regression splines and the SPDE approach.

@@

The predicted mean of the posterior intensity is shown in \autoref{fig:intensity-mean-cv} along with a map of the coefficient of variation (CV) of the posterior intensity field on a regular prediction grid with cell area of approximately 1.7 hectares.  
\begin{figure}[h]
	\begin{center}
		\includegraphics[scale=0.525]{figures/intensity_mean_cv_sd.png}
		\caption{The A) mean, B) CV and C) SD of the posterior intensity field.  Units are per m$^2$}
		\label{fig:intensity-mean-cv}
	\end{center}
\end{figure}
This shows a region of high intensity in the south and much lower intensity in the north, agreeing with a standard two-stage analysis of the same data (see Supplementary material).  The CV plot shows that the CV in the posterior intensity field is lower in areas with greater sampling effort.  There is a clear indication of preferential sampling with more survey effort in the south, where densities are higher, compared to the north  (see \autoref{fig:2002studyareapointspt}).  In the north, estimated intensities are lower, which also contributes to larger CV values.  The standard deviation map shows stronger overall posterior variability in the south where intensity is larger.  We also map the lower and upper quantiles corresponding to a 95\% credible interval for each prediction location.  [[Say something about these - hard to know what to say about them?]]   (\autoref{fig:intensity-quantiles}).
\begin{figure}[h]
	\begin{center}
		\includegraphics[scale=0.525]{figures/intensity_quantiles.png}
		\caption{The predicted posterior intensity quantiles:  A) 0.025 quantile B) 0.975 quantile.  Both plots plots use the same colour scale}
		\label{fig:intensity-quantiles}
	\end{center}
\end{figure}

@@

We plot the posterior abundance estimate using a Monte Carlo method.
\begin{figure}[h]
	\begin{center}
		\includegraphics[scale=0.525]{figures/realized_abundance_posterior.png}
		\caption{Posterior abundance}
		\label{fig:realized-abundance-posterior}
	\end{center}
\end{figure}
Let $n$ denote the abundance of the population within the region of interest and $N = \int_{\Omega}\lambda(s)\mathrm{d}s$ its corresponding random variable within our Bayesian framework.  Integrating the mean of the posterior intensity field provides a point estimate for the abundance, i.e. the expected abundance.  We approximate the posterior $\pi(N | \bm{Y})$ by a Monte Carlo method.  Taking $m$ Monte Carlo samples  $\lambda^{(1)}, \ldots, \lambda^{(m)}$ of the posterior intensity field we estimate the posterior for the abundance as $\pi(N | \bm{Y}) \approx 1 / m \sum_{i=1}^m \pi (N | \lambda = \lambda^{(i)})$. Each $\pi(N | \lambda = \lambda_i)$ is a Poisson probability mass function with rate parameter $\int_{\Omega}\lambda_i(s)\mathrm{d}s$. \autoref{fig:realized-abundance-posterior} shows the approximate posterior for $N$ with $m = 20,000$.  This allows us to estimate the probability for any specific value of realised abundance $n$ by estimated $\pi(N = n | \lambda)$.  Strictly speaking

@@

There are several natural extensions to the model.  For this analysis we only considered a single survey year.  Spatio-temporal extensions to this are possible, for example by considering a space-time interaction models.  \texttt{R-INLA} has the ability to fit space-time effects that can be represented as a Kronecker product of the relevant precision matrices \citep{blangiardo_spatial_2013, yuan_point_2017}.  An example of this type of space-time interaction effect is an interaction between the spatial SPDE with an auto-regressive temporal process.  \cite{blangiardo_spatial_2013} show how this can be done with numerous examples and code snippets.
